\documentclass[11pt]{article}

% Change "review" to "final" to generate the final (sometimes called camera-ready) version.
% Change to "preprint" to generate a non-anonymous version with page numbers.
\usepackage[preprint]{acl}

% Standard package includes
\usepackage{times}
\usepackage{latexsym}

% For proper rendering and hyphenation of words containing Latin characters (including in bib files)
\usepackage[T1]{fontenc}
% For Vietnamese characters
% \usepackage[T5]{fontenc}
% See https://www.latex-project.org/help/documentation/encguide.pdf for other character sets

% This assumes your files are encoded as UTF8
\usepackage[utf8]{inputenc}

% This is not strictly necessary, and may be commented out,
% but it will improve the layout of the manuscript,
% and will typically save some space.
\usepackage{microtype}

% This is also not strictly necessary, and may be commented out.
% However, it will improve the aesthetics of text in
% the typewriter font.
% \usepackage{inconsolata}

%Including images in your LaTeX document requires adding
%additional package(s)
\usepackage{graphicx}

% If the title and author information does not fit in the area allocated, uncomment the following
%
%\setlength\titlebox{<dim>}
%
% and set <dim> to something 5cm or larger.

\title{An Empirical Analysis of Hallucination Behavior in LLMs in the Medical Domain}

\author{Aidan Perez \\
  MS in AI, Boston University\\
  \texttt{aperez00@bu.edu} \\\And
  Dang Dinh \\
  MS in Data Science, Boston University\\
  \texttt{dddinh@bu.edu} \\\AND
  Kaveri Goel \\
  MS in AI, Boston University\\
  \texttt{kvrgl@bu.edu} \\\And
  Ying Huang \\
  MS in CS, Boston University\\
  \texttt{yinghy@bu.edu} }

\begin{document}
\maketitle
\begin{abstract}
Large Language Models (LLMs) have demonstrated remarkable capabilities in generating fluent, contextually relevant text. However, they frequently produce hallucinations — outputs that appear plausible but are factually incorrect, unsupported, or fabricated. This problem is especially critical in high-stakes domains such as medicine, where incorrect information can have serious real-world consequences. While hallucination has been studied in general NLP settings, relatively little empirical work systematically characterizes hallucination behavior specifically within the medical domain — where ground truth is verifiable, the cost of error is high, and LLMs are increasingly being adopted (e.g., in clinical decision support and patient-facing chatbots). 
In this paper, we present a comprehensive empirical analysis of hallucination behavior in LLMs within the medical domain. We evaluate multiple state-of-the-art LLMs on a diverse set of medical tasks, including question answering, summarization, and clinical note generation. We systematically categorize the types of hallucinations observed, analyze their frequency and severity, and identify common patterns and triggers. Our findings provide critical insights into the nature of hallucination in medical LLMs, highlight the risks associated with their deployment in clinical settings, and suggest directions for future research aimed at mitigating hallucination and improving the reliability of LLMs in healthcare applications.
\end{abstract}
\section{Introduction}
Large Language Models (LLMs) have revolutionized natural language processing (NLP) by demonstrating unprecedented capabilities in generating coherent, contextually relevant text across a wide range of tasks. However, a significant challenge that has emerged with the deployment of LLMs is their propensity to produce hallucinations — outputs that may be fluent and plausible but are factually incorrect, unsupported by evidence, or entirely fabricated. This issue is particularly concerning in high-stakes domains such as medicine, where the dissemination of incorrect information can lead to serious consequences for patient care and safety. Despite the growing body of research on hallucination in general NLP contexts, there remains a paucity of systematic empirical analyses that specifically characterize hallucination behavior in medical LLMs. Given the increasing adoption of
LLMs in healthcare applications, such as clinical decision support systems and patient-facing chatbots, it is imperative to understand the nature and implications of hallucination in this domain. In this paper, we present a comprehensive empirical analysis of hallucination behavior in LLMs within the medical domain. We evaluate multiple state-of-the-art LLMs across a diverse set of medical tasks, including question answering, summarization, and clinical note generation. Our analysis categorizes the types of hallucinations observed, quantifies their frequency and severity, and identifies common patterns and triggers. The insights gained from our study provide critical guidance for the responsible deployment of LLMs in healthcare settings and inform future research aimed at mitigating hallucination and enhancing the reliability of medical LLMs. Solving this problem is crucial 
for ensuring that LLMs can be safely and effectively integrated into clinical workflows, ultimately improving patient outcomes and advancing the field of medical AI.  This is a hard problem to solve, and we hope that our work will inspire further research in this area.  The reason that this problem is so hard to solve is that it requires a deep understanding of the underlying language model, as well as the ability to identify and mitigate hallucinations in a way that does not compromise the model's ability to generate useful and relevant text.  Additionally, the medical domain is particularly challenging due to the complexity and nuance of medical language, as well as the high stakes involved in providing accurate information.  We hope that our work will provide a foundation for future research in this area, and that it will ultimately lead to the development of more reliable and trustworthy LLMs for use in healthcare applications.
\section{Progress} 
\subsection{Methods}
% TODO: Cite MedQA and PubMedQA datasets
% We might want to list this step by step to make it clearer
To conduct our empirical analysis of hallucination behavior in medical LLMs, we have selected two widely used datasets: MedQA and PubMedQA. The MedQA dataset consists of multiple-choice questions derived from the United States Medical Licensing Examination (USMLE), covering a broad range of medical topics and requiring complex reasoning and domain-specific knowledge. The PubMedQA dataset, on the other hand, consists of yes/no questions based on abstracts from the PubMed database, which require the model to extract relevant information from the provided abstracts to generate accurate responses. We have chosen these datasets because they represent different types of medical tasks and require varying levels of reasoning and knowledge, allowing us to comprehensively evaluate the hallucination behavior of LLMs across a spectrum of medical applications. We have implemented a systematic evaluation framework that includes regular prompting, chain-of-thought prompting, and Retrieval-Augmented Generation (RAG)
to assess the performance of the LLMs on these datasets. Regular prompting involves providing the model with a straightforward prompt to generate a response, while chain-of-thought prompting encourages the model to generate intermediate reasoning steps before arriving at a final answer. RAG combines retrieval-based methods with generative models, allowing the model to access external knowledge sources to generate more accurate and relevant responses. By comparing the outputs generated by the LLMs under these different prompting strategies, we aim to identify patterns of hallucination and evaluate the effectiveness of various techniques for mitigating this issue in medical LLMs.\\
\subsection{Preliminary Findings}

We conducted a preliminary comparison of Meta-Llama-3-8B-Instruct under two inference conditions on the MedQA benchmark: standard zero-shot prompting and chain-of-thought (CoT) prompting. On 500 evaluation samples, zero-shot prompting yielded 302 correct predictions (60.4\% accuracy), a hallucination rate of 39.6\%, 4 parse failures, and an average BERTScore F1 of 0.819.
In contrast, CoT prompting yielded 282 correct predictions (56.4\% accuracy), a hallucination rate of 43.6\%, 18 parse failures, and an average BERTScore F1 of 0.814. These results indicate that, in our current MedQA setting, CoT does not outperform the zero-shot baseline for Llama3. Instead, it is associated with slightly lower accuracy, a higher hallucination rate, and reduced output stability.

Inspection of the raw generations suggests that this pattern may be partly explained by differences in output format between the two prompting conditions. Under zero-shot prompting, Llama3 usually provides an answer letter that can be extracted reliably, although many responses still include additional explanatory text. 
Under CoT prompting, the model more frequently produces longer reasoning traces and explicit answer markers, but these generations are also more likely to contain repetition, trailing phrases, and other formatting artifacts. As a result, CoT outputs are more vulnerable to parsing errors in the current pipeline.

Taken together, these findings suggest that explicit reasoning prompts do not automatically improve performance on medical multiple-choice QA. In our current setup, longer CoT generations may create additional opportunities for noisy intermediate reasoning and answer extraction failure. 
We therefore treat this as a preliminary result tied to the current prompting and evaluation design, rather than as a general claim that CoT is ineffective in medical QA. In future work, we plan to investigate whether stricter output constraints and more robust answer extraction can improve the reliability of CoT prompting.
\subsection{RAG Preliminary Findings}

We evaluated three models under Retrieval-Augmented Generation (RAG) conditions on the PubMedQA benchmark using BM25 retrieval with top-2 abstracts. On 450 evaluation samples, Meta-Llama-3-8B-Instruct achieved 73.4\% accuracy with a hallucination rate of 28.9\% and an average BERTScore F1 of 0.850. BioMistral-7B achieved 52.9\% accuracy with a hallucination rate of 54.9\% and BERTScore F1 of 0.816. BioGPT achieved 54.4\% accuracy but with a hallucination rate of 89.1\% and a parse failure rate of 80\%, indicating the model struggled to generate structured yes/no/maybe responses.

These preliminary findings suggest that general-purpose models such as Llama3 benefit more from RAG than smaller domain-specific models. BioGPT's high parse failure rate indicates that the current prompting strategy may not be well-suited for smaller generative models, which we plan to address in future experiments. Comparison between RAG and zero-shot conditions on the same PubMedQA questions remains pending and will be included in the final report.
\end{document}